---
title: "hw02_gapminder"
output: github_document
---

# Homework 02: Explore Gapminder and use dplyr

## Bring rectangular data in

```{r}
library(gapminder)
library(tidyverse)
```

## Smell test the data

Explore the `gapminder` object:

* Is it a data.frame, a matrix, a vector, a list?
* What is its class?
* How many variables/columns?
* How many rows/observations?


```{r}
str(gapminder)
```

*The output shows that the gapminder is a data frame (Classes are ‘tbl_df’, ‘tbl’ and 'data.frame'). There are 6 columns/variables and 1704 rows/observations.*

* Can you get these facts about "extent" or "size" in more than one way? Can you imagine different functions being useful in different contexts?

```{r}
nrow(gapminder)
ncol(gapminder)
```

*Using `ncol` and `nrow` is another way to get these dimensions. While `str` gives us more information, these 2 can be useful, especially if you want to use just the number of rows/cols. One might do so to calculate mean or loop through columns.*

* What data type is each variable?

From `str`, we can see the type for each variable.

* country is Factor
* continent is Factor
* year is int
* lifeExp is num
* pop is int
* gdpPercap is num


## Explore individual variables

Pick __at least__ one categorical variable and at least one quantitative variable to explore.

* What are possible values (or range, whichever is appropriate) of each variable?
* What values are typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.
* Feel free to use summary stats, tables, figures. We're NOT expecting high production value (yet).

*We can see a summary with of all the variable using `summary`.*

```{r}
summary(gapminder)
```

*I have chosen to explore gdpPercap and continent*

### Exploring "gdpPercap"

*I can see a summary of gdpPercap specifically that contains its min, max, mean, median, 1st quartile and 3rd quartile.*
```{r}
summary(gapminder$gdpPercap)
```

*Alternatively, I can calculate several of these individually.*
```{r}
min(gapminder$gdpPercap)
max(gapminder$gdpPercap)
mean(gapminder$gdpPercap)
median(gapminder$gdpPercap)
```

*I can get a sense of the distribution of the data by looking at its variance or standard deviation.*

```{r}
var(gapminder$gdpPercap)
sd(gapminder$gdpPercap)
```

### Exploring "continent"

*Since continent is a factor, let's look at its levels.*

```{r}
levels(gapminder$continent)
```

*There are 5 different levels. We can count them with `summary` as well to see how many of the data points fall in each continent.*

```{r}
summary(gapminder$continent)
```

## Explore various plot types

Make a few plots, probably of the same variable you chose to characterize numerically. You can use the plot types we went over in class (cm006) to get an idea of what you'd like to make. Try to explore more than one plot type. **Just as an example** of what I mean:

* A scatterplot of two quantitative variables.
* A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.
* A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.

You don't have to use all the data in every plot! It's fine to filter down to one country or small handful of countries.

## Use `filter()`, `select()` and `%>%`

Use `filter()` to create data subsets that you want to plot.

Practice piping together `filter()` and `select()`. Possibly even piping into `ggplot()`.

## Alvin's Exploration Plots

### Life Expectancy and Years

```{r}
gapminder %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_line(lwd = 0.5, show.legend = FALSE) + 
  facet_wrap(~ continent) +
  theme_bw()
```

*Here, I plot the lifeExp of each country as a different colour of the years. Over time, life expenctancy increases. I decided to pipe in the gapminder dataset with `%>%`. Additional things I did were to hide the legend, change the line width, change to bw theme, and facet by the different continents.*

### GDP per Capita and continent

```{r}
ggplot(gapminder, aes(x = continent, y = gdpPercap)) +
  scale_y_log10() +
  theme_bw() +
  geom_violin(aes(fill = continent), show.legend = FALSE) +
  geom_jitter(alpha = 0.2)
```

*Now, I want to look at the distrubition of GDP per capita for each continent. There is a violin plot, with different fill for each continent. Overlaid is a jitter plot for each point to better visualize distributaion. I've chosen a log Y scale for the GDP per cap. We can see the African countries are amongst the poorest. However, there is a skew, with a few rather wealthier African countries. Given the log scale, this highlights great disparity. Simila observation with Asian countries, showing a wide spread.*

## Life Expectancy and GDP per capita

```{r}
gapminder %>%
  filter(continent == "Asia") %>%
  ggplot(aes(x = lifeExp, y = gdpPercap, colour = year)) +
  geom_point() +
  scale_y_log10() +
  theme_bw() +
  geom_smooth(method = "lm") +
  scale_colour_gradient(low = "white", high = "darkblue")
```

*This time, I `filter` for Asian countries. I look at the relationship between lifeExp and gdpPercap, which is a postive trend (demonstrated with the geom_smooth). The GDP is on a log Y scale again. Additionally, I coloured the points, by year. We can see that the more recent years have a tread of higher lifeExp and higher GDP.*

## But I want to do more!

### More with `dplyr`

I want to demonstrate the `mutate` function. 

```{r}
gapminder %>%
  select(country, year, pop, gdpPercap) %>%
  mutate(GDP = pop * gdpPercap)
```

*First, I take gapminder and using pipes, I select for a few relevant columns. GDP per capita is GDP divided by population. I want a new column with GDP. Using `dplyr::mutate`, I make a new column, GDP, which is the total pop X gdpPercap.*

### Even more

*For people who want to take things further.*

Evaluate this code and describe the result. Presumably the analyst's intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```
This is not correct. Note there is missing data.

```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))

filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```
Here are 2 possible solution. One uses `|` as an OR statement. The other uses `%in%`.